---
title: "TP Ecologie Biodiversité Montmaur"
author: "Eric Marcon d'après Bastien Mérigot et Guillaume Papuga"
date: "`r Sys.Date()`"
bibliography: references.bib
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
# Installation of packages if necessary
InstallPackages <- function(Packages) {
  InstallPackage <- function(Package) {
    if (!Package %in% installed.packages()[, 1]) {
      install.packages(Package, repos="https://cran.rstudio.com/")
    }
  }
  invisible(sapply(Packages, InstallPackage))
}
# Basic packages
InstallPackages(c("bookdown", "formatR", "tidyverse"))
# Specific packages
InstallPackages(
  c(
    "ade4", 
    "vegan"
  )
)
# knitr options
knitr::opts_chunk$set(echo = TRUE, cache = FALSE, message = FALSE, warning = FALSE)
# Load the tidyverse
library("tidyverse")
```

# Data

## Base R version 

Load the data:

```{r}
montmaur <- read.csv2("data/species.csv")
```

Interactive version:

```{r, eval=FALSE}
montmaur <- read.table(file.choose(), sep=";", dec=",", header=T)
# Edit the data
montmaur <- edit(montmaur) 
```

Editing the data interactively is a bad practice because it is not reproducible.
Do rather edit the original file.

Explore the data:
```{r}
head(montmaur)
str(montmaur)
summary(montmaur)
```

Map the plots:

```{r}
# Force the aspect ratio
plot(x = montmaur$longitude, y = montmaur$longitude, asp = 1)
```
Avoid repeating `montmaur$`:

- With `attach()`: bad practice.
See `?attach`.
```{r}
attach(montmaur)
plot(x = longitude, y = longitude, asp = 1)
# Don't forget to detach after use!
detach(montmaur)
```

- With `with()`.
```{r}
with(montmaur,
  # Indent the code
  plot(x = longitude, y = longitude, asp = 1)
  # Close the bracket
)
```


Are some plots empty?
```{r}
# The first 10 columns contain plot data, the others, the species abundances
min(rowSums(montmaur[, -(1:10)]))
```
Are some species missing?
```{r}
# The first 10 columns contain plot data, the others, the species abundances
which(colSums(montmaur[, -(1:10)]) == 0)
# Numbers are those of the columns. They are named.
```
Create a dataframe with species abundances only, and eliminate absent species

```{r}
montmaur_sp <- montmaur[, -(1:10)][, -which(colSums(montmaur[, -(1:10)]) == 0)]
# Add the plot names
rownames(montmaur_sp) <- montmaur$code_rel
```

Make dataframes for the north and south face only:
```{r}
montmaur_sp_n <- montmaur_sp[montmaur$versant == "N", ]
montmaur_sp_s <- montmaur_sp[montmaur$versant == "S", ]
```

Graphical representation of the data table:

```{r}
library("ade4")
table.value(montmaur_sp_n, clegend=0) 
```

## Tidyverse

Read the data.

```{r}
library("tidyverse")
montmaur <- read_csv2("data/species.csv")
```

Make a pipeline to select rows and columns:

```{r}
montmaur %>%
  # Filter the north face
  filter(versant == "N") %>% 
  # Select species columns
  select(!(Groupe:rec_tot)) %>% 
  # Eliminate absent species
  select(where(~sum(.) != 0))
```


# Accumulation curves

## Assess diversity at a sampling level

```{r}
library("vegan")
sac <- specaccum(montmaur_sp)
plot(sac)
sac_n <- specaccum(montmaur_sp_n) 
sac_s <- specaccum(montmaur_sp_s)
```

The number of species is in the `$richness` item of the `specaccum` object, which is a list.

```{r}
sac$richness
```

Plot north and south SAC

```{r}
plot(sac_s, ci.type = "line", ci.col = "red")
plot(sac_n, ci.type = "line", ci.col = "blue", add = TRUE)
```

## Rarefaction to compare diversity

Rarefy the data to the smallest sample size.

```{r}
# Find the smallest number of plots
rarefied_n <- min(nrow(montmaur_sp_n), nrow(montmaur_sp_s))
# Estimate and estimator sd 
# Rarified richness at n=17 quadra
sac_n$richness[rarefied_n]
sac_n$sd[rarefied_n]
sac_s$richness[rarefied_n]
sac_s$sd[rarefied_n]
```

Assuming normality of the rarefied richness, conclude wether the north or south face is more diverse.


# Rank-abundance diagram

Compute the probability for an area to be covered by each species.

```{r}
# Sum the species cover
sp_cover <- colSums(montmaur_sp)
# Normalise
sp_cover <- sp_cover / sum(sp_cover)
```

Try to fit the distribution to a well-known one.

```{r}
radfit(sp_cover)
plot(radfit(sp_cover))
```

# Taxonomic Diversity

```{r}
# Specific richness S
sp_richness <- specnumber(montmaur_sp, MARGIN = 1)
```


```{r}
# Species occurrences
sp_occurrence <- specnumber(montmaur_sp, MARGIN = 2)
plot(radfit(sp_occurrence))
```
## Shannon, Simpson, entropy


## Diversity


# Functional diversity


# Beta diversity

## Definitions

## Differentiation

## Partitioning

