---
title: "TP Ecologie Biodiversité Montmaur"
author: "Eric Marcon d'après Bastien Mérigot et Guillaume Papuga"
date: "`r Sys.Date()`"
bibliography: [references.bib, packages.bib]
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
# Installation of packages if necessary
InstallPackages <- function(Packages) {
  InstallPackage <- function(Package) {
    if (!Package %in% installed.packages()[, 1]) {
      install.packages(Package, repos="https://cran.rstudio.com/")
    }
  }
  invisible(sapply(Packages, InstallPackage))
}
# Basic packages
InstallPackages(c("bookdown", "formatR", "tidyverse"))
# Specific packages
packages <- c(
  "ade4", 
  "adiv",
  "entropart",
  "FD",
  "GoFKernel",
  "vegan"
)
InstallPackages(packages)
# Create packages.bib file to reference them as @R-package
knitr::write_bib(packages, file = "packages.bib")
# knitr options
knitr::opts_chunk$set(echo = TRUE, cache = FALSE, message = FALSE, warning = FALSE)
# Load the tidyverse
library("tidyverse")
```

# Data

The classical shape of biodiversity data is a matrix where columns are species and rows are plots.
Values are abundances.

## Base R version 

Load the data:

```{r}
montmaur <- read.csv2("data/species.csv")
```

Interactive version:

```{r, eval=FALSE}
montmaur <- read.table(file.choose(), sep=";", dec=",", header=T)
# Edit the data
montmaur <- edit(montmaur) 
```

Editing the data interactively is a bad practice because it is not reproducible.
Do rather edit the original file.

Explore the data:
```{r, eval=FALSE}
head(montmaur)
str(montmaur)
summary(montmaur)
```

Map the plots:

```{r}
# Force the aspect ratio
plot(x = montmaur$longitude, y = montmaur$longitude, asp = 1)
```

Avoid repeating `montmaur$`:

- With `attach()`: bad practice.
See `?attach`.
```{r}
attach(montmaur)
plot(x = longitude, y = longitude, asp = 1)
# Don't forget to detach after use!
detach(montmaur)
```

- With `with()`.
```{r}
with(montmaur,
  # Indent the code
  plot(x = longitude, y = longitude, asp = 1)
  # Close the bracket
)
```


Are some plots empty?

```{r}
# The first 10 columns contain plot data, the others, the species abundances
min(rowSums(montmaur[, -(1:10)]))
```
Are some species missing?

```{r}
# The first 10 columns contain plot data, the others, the species abundances
which(colSums(montmaur[, -(1:10)]) == 0)
# Numbers are those of the columns. They are named.
```
Create a dataframe with species abundances only, and eliminate absent species

```{r}
montmaur_sp <- montmaur[, -(1:10)][, which(colSums(montmaur[, -(1:10)]) > 0)]
# Add the plot names
rownames(montmaur_sp) <- montmaur$code_rel
```

Make dataframes for the north and south face only:
```{r}
montmaur_sp_n <- montmaur_sp[montmaur$versant == "N", ]
montmaur_sp_s <- montmaur_sp[montmaur$versant == "S", ]
```

Graphical representation of the data table:

```{r}
library("ade4")
table.value(montmaur_sp_n, clegend=0) 
```

## Tidyverse

Read the data.

```{r}
library("tidyverse")
montmaur <- read_csv2("data/species.csv")
```

Make a pipeline to select rows and columns:

```{r}
montmaur %>%
  # Filter the north face
  filter(versant == "N") %>% 
  # Select species columns
  select(!(Groupe:rec_tot)) %>% 
  # Eliminate absent species
  select(where(~sum(.) != 0))
```


# Rank-abundance diagram

Course: @Marcon2014d, section [2.4](https://ericmarcon.github.io/MesuresBioDiv2/chap-Outils.html#distribution-de-labondance-des-esp%C3%A8ces-sad).

## Abundances: species cover

Compute the probability for an area to be covered by each species.

```{r}
# Sum the species cover
montmaur_cover <- colSums(montmaur_sp)
# Normalise
montmaur_cover <- montmaur_cover / sum(montmaur_cover)
```

Try to fit the distribution to a well-known one.

```{r}
library("vegan")
radfit(montmaur_cover)
plot(radfit(montmaur_cover))
```

## Occurrence

Count the number of plots where each species is observed.

```{r}
# Species occurrences
montmaur_occurrence <- specnumber(montmaur_sp, MARGIN = 2)
radfit(montmaur_occurrence)
plot(radfit(montmaur_occurrence))
```

## Compare abundance and occurence

```{r}
plot(x = montmaur_cover, y = montmaur_occurrence)
```



# Taxonomic Diversity

Course: @Marcon2014d, chapters [3](https://ericmarcon.github.io/MesuresBioDiv2/chap-MesuresNeutres.html) and [4](https://ericmarcon.github.io/MesuresBioDiv2/entropie.html).

## Richness

Count the number of species per plot.

```{r}
# Specific richness S
(plot_richness <- specnumber(montmaur_sp, MARGIN = 1))
```

## Accumulation curves

The level of diversity depends on the sampling effort.

### Assess diversity at a sampling level

Course: @Marcon2014d, sections [2.1](https://ericmarcon.github.io/MesuresBioDiv2/chap-Outils.html#courbes-daccumulation) and 2.2.

Accumulation curves can be built by package **vegan** [@R-vegan].

```{r}
library("vegan")
sac <- specaccum(montmaur_sp)
plot(sac)
sac_n <- specaccum(montmaur_sp_n) 
sac_s <- specaccum(montmaur_sp_s)
```

The number of species is in the `$richness` item of the `specaccum` object, which is a list.

```{r}
sac$richness
# Total richness is also
sum(colSums(montmaur_sp) != 0)
```

Plot north and south SAC:

```{r}
plot(sac_s, ci.type = "line", ci.col = "red")
plot(sac_n, ci.type = "line", ci.col = "blue", add = TRUE)
```

### Rarefaction to compare diversity

Course: @Marcon2014d, chapter [21](https://ericmarcon.github.io/MesuresBioDiv2/chap-Accumulation.html#chap-Accumulation).

Rarefy the data to the smallest sample size.

```{r}
# Find the smallest number of plots
rarefied_n <- min(nrow(montmaur_sp_n), nrow(montmaur_sp_s))
# Estimate and estimator sd 
# Ratified richness at n=rarefied_n plots
sac_n$richness[rarefied_n]
sac_n$sd[rarefied_n]
sac_s$richness[rarefied_n]
sac_s$sd[rarefied_n]
```
Compare the diversity of north and south face.


## Extrapolation

Extrapolation is possible, up to twice the sample size without too much uncertainty.
See the **iNext** package [@Hsieh2014] for the [framework](https://cran.r-project.org/web/packages/iNEXT/vignettes/Introduction.pdf).


```{r}
library("entropart")
Richness(montmaur_occurrence, Level = 2 * sum(montmaur_occurrence))
```
Pot the accumulation curve:

```{r}
autoplot(
  DivAC(
    montmaur_occurrence,
    q = 0, 
    n.seq = 1:(2 * sum(montmaur_occurrence))
  )
)
```


## Asymptotic richness

Course: @Marcon2014d, sections [2.2](https://ericmarcon.github.io/MesuresBioDiv2/chap-Outils.html#diversit%C3%A9-asymptotique) and [3.1](https://ericmarcon.github.io/MesuresBioDiv2/chap-MesuresNeutres.html#sec-Richesse)

Requires estimators based on the number of individuals or on the number of occurrences, available in package **entropart** [@R-entropart].

```{r}
Richness(montmaur_occurrence, Correction = "Chao1")
```

## Shannon, Simpson, entropy

Richness minus one species, Shannon [-@Shannon1948] and Simpson [-@Simpson1949] indices measure entropy, i.e. an amount of information.

In a community:

- The _probability_ $p_s$ of a species is the probability for an individual to belong to it.
When individuals are not well-defined, another definition must be chosen: with abundance-dominance data, it is the proportion of total coverage occupied by the species.
- The _rarity_ of a species is the reciprocal of its probability, i.e. $1/p_s$.
- The information brought by the species is:
    - $(1-p_s)/p_s$ for richness (full line) 
    - the logarithm of rarity for Shannon ($\ln(1/p_s)$) (long dashes), 
    - $1-p_s$ for Simpson (sort dashes).

```{r ICode}
I0 <- function(p) (1 - p) / p
I1 <- function(p) -log(p)
I2 <- function(p) 1 - p
ggplot(data.frame(x=c(0, 1)), aes(x)) + 
    stat_function(fun = I0) +
    stat_function(fun = I1, lty = 2) +
    stat_function(fun = I2, lty = 3) +
    coord_cartesian(ylim = c(0, 10)) +
    labs(x = "p", y = "I(p)")
```

- The entropy of the community is the average information brought by its species.

The information function gives more (richness) or less (Simpson) weight to rare species.

```{r}
# Shannon
Shannon(montmaur_occurrence, Correction = "None")
# Asymptotic
Shannon(montmaur_occurrence)
# Cover data
Shannon(montmaur_cover)
```

Simpson's entropy is the probability for two individuals to belong to different species.

```{r}
# Simpson
Simpson(montmaur_occurrence, Correction = "None")
# Asymptotic
Simpson(montmaur_occurrence)
# Cover data
Simpson(montmaur_cover)
```
The generalized entropy extends this approach.
The deformed logarithm function of order $q$ is $\ln_q(x) = \frac{x^{1-q} - 1}{1 - q}$.

- the natural logarithm for $q=1$ : $\ln_1(x) = \ln(x)$.
- a more concave function for $q > 1$ : $\ln_2(x) = 1 - 1/x$.
- a less concave function for $q < 1$ : $\ln_0(x) = x - 1$.

Then:

- entropy of order q is the average deformed logarithm (of order q) of rarity: $^{q}H = \sum_s {\ln_q(1/p_s)}$.
- entropy of order 0 is richness minus 1.
- entropy of order 1 is Shannon's entropy.
- entropy of order 2 is Simson's entropy.
- the greater $q$, the less importance is given to rare species.

## Diversity

Diversity sensu stricto is the number of equiprobable species that would yield the same entropy as the data.
It is called a Hill [-@Hill1973] number or a number equivalent or the effective number of species [@Jost2006].

It is the exponential of Shannon's entropy and the reciprocal of Simpson's similarity.

```{r}
# Shannon
Diversity(montmaur_cover, q = 1)
exp(Shannon(montmaur_cover))
# Simpson
Diversity(montmaur_cover, q = 2)
1 / (1 - Simpson(montmaur_cover))
```

In general, diversity of order $q$ is the deformed exponential of entropy of order $q$.

```{r}
# Simpson
expq(Simpson(montmaur_cover), q = 2)
# From diversity to entropy
lnq(Diversity(montmaur_cover, q = 2), q = 2)
```

## Diversity profile

Plot diversity with respect to its order.

```{r}
autoplot(CommunityProfile(Diversity, montmaur_cover))
```


# Functional diversity

```{r}
traits <- read.csv2("data/traits.csv")
# chlorophyll
trait_chlorophyll <- aggregate(chlorophylle ~ code_espece, data = traits, FUN = mean)
# Biological type
trait_type <- aggregate(type_biologique ~ code_espece, data = traits, FUN = unique)
# Trait table
traits <- cbind(trait_chlorophyll, trait_type)
# Delete doubled column and keep observed species only
traits <- traits[traits$code_espece %in% colnames(montmaur_sp), -3]
head(traits)
```

## Community weighted mean

Course: @Marcon2014d, section [7.7](https://ericmarcon.github.io/MesuresBioDiv2/mesures-particuli%C3%A8res.html#diversit%C3%A9-et-moyenne).

Defined by @Garnier2004.
CWM is a mean, not a diversity.

```{r}
species_weights <- colSums(montmaur_sp)[colnames(montmaur_sp) %in% traits$code_espece]
mean(traits$chlorophylle, w = species_weights)
# By type (unweighted)
aggregate(chlorophylle ~ type_biologique, data = traits, FUN = mean)
```

For intensive use, see `BAT::cwm()`.

## Functional Distance

Course: @Marcon2014d, chapter [6](https://ericmarcon.github.io/MesuresBioDiv2/chap-cadrephyfonc.html#chap-cadrephyfonc).

Functional distance between species following @Gower1971.

```{r}
library("FD")
# Distances based on traits, not including species names.
traits_distance <- gowdis(traits[, -1])
plot(density(traits_distance))
```

## Quadratic entropy

Course: @Marcon2014d, section [7.6](https://ericmarcon.github.io/MesuresBioDiv2/mesures-particuli%C3%A8res.html#sec-Rao).

Rao's entropy (aka quadratic entropy) is the average distance between two individuals.

```{r, eval=FALSE}
library("adiv")
montmaur_qe <- QE(
  comm = montmaur_sp[, traits$code_espece], 
  dis = traits_distance
)
```


## Hierarchical classification

Functional classification of species (complementarity-redundancy) with CAH (mean linkage) based on Gower distances:

```{r}
# Restore the species names
attr(traits_distance, "Labels") <- traits$code_espece
species_fhc <- hclust(traits_distance, method = "average")
plot(species_fhc, hang = -1)
```


# Phylogenetic diversity

Course: @Marcon2014d, chapter [8](https://ericmarcon.github.io/MesuresBioDiv2/chap-Phyloentropie.html#chap-Phyloentropie).

Defined to assess the dissimilarities between species summarized in a dendrogram.
We'll use the functional dendrogram above as an exercice, even though it is not a good approach to measure functional diversity.

The dendrogram is cut into slices at each node height.
Entropy is calculated in each slice and averaged along the tree: it is the average information along time (if the dendrogram represents evolutionary times) and diversity is its exponential.

PD [@Faith1992], aka FD [@Petchey2006] is the total length of branches, i.e. richness.

```{r}
# Subset the data: keep species of the tree only
montmaur_reduced <- montmaur_occurrence[
  names(montmaur_occurrence) %in% species_fhc$labels
]
# FD
PDFD(as.ProbaVector(montmaur_reduced), species_fhc)
```
Entropy of order 2 is the average distance between two individuals, i.e. Rao's entropy.

```{r}
# Asymptotic Rao's entropy in a normalized tree
PhyloEntropy(montmaur_reduced, q = 2, Tree = species_fhc, Normalize = TRUE)
```

Diversity is its deformed exponential of order 2, i.e. 1/(1 - entropy).

```{r}
PhyloDiversity(montmaur_reduced, q = 2, Tree = species_fhc, Normalize = TRUE)$Total
```

# Beta diversity

Course: @Marcon2014d, chapter [10](https://ericmarcon.github.io/MesuresBioDiv2/sec-betaCadre.html).

## Partitioning

Course: @Marcon2014d, chapter [12](https://ericmarcon.github.io/MesuresBioDiv2/chap-DedompHCDT.html#chap-DedompHCDT).


Défined by @Whittaker1960. 
Multiplicative partitioning: 

```{r}
# Total number of species
richness_gamma <- sum(colSums(montmaur_sp) > 0)
# Average number of species per plot
richness_plot <- apply(montmaur_sp, MARGIN = 1, FUN = function(plot) sum(plot > 0))
richness_alpha <- mean(richness_plot)
richness_beta <- richness_gamma / richness_alpha
```

$\beta$ diversity is the effective number of plots, i.e. the number of equally-weighted plots that would yield the same diversity as the data.

Additive partitioning [@Lande1996]:

```{r}
richness_gamma - richness_alpha
```

Additive partitioning is valid for entropy.


## Differentiation

Course: @Marcon2014d, chapter [11](https://ericmarcon.github.io/MesuresBioDiv2/chap-BetaPaires.html#chap-BetaPaires).

How different are two plots?

Jaccard dissimilarity is based on occurrence data.

```{r}
library("vegan")
jaccard <-vegdist(montmaur_sp, method = "jaccard")
hist(jaccard)
# Smooth the histogram, no values above 1
library("GoFKernel")
plot(density.reflected(jaccard, upper = 1))
```

Bray-Curtis dissimilarity is based on abundance data.

```{r}
bray <-vegdist(montmaur_sp, method = "bray")
plot(density.reflected(bray, upper = 1))
```


## Conciliation

The figure of plots after a non-scaled, centred PCA shows the contribution of each plot to Simpson's $\beta$ entropy (partitioning).
The distance between plots is Simpson's $\beta$ entropy between them (differentiation).

```{r}
montmaur_pca <- dudi.pca(
  montmaur_sp/rowSums(montmaur_sp), 
  scannf = FALSE, 
  nf = 2,
  center = TRUE,
  scale = FALSE
)
scatter(montmaur_pca, clab.col = 0)
```


# References
